---
title: "SiteSelectR_Intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SiteSelectR_Intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Monitoring networks are crucial for understanding how ecosystems respond to environmental change. Apparent trends in biodiversity can be driven by site selection bias towards easily accessed sites, not necessarily the most representative of the landscape – future inference could misrepresent trends due to the original network design. SiteSelectR is an r package aimed at creating a framework for long-term ecological monitoring that decreases spatial and environmental bias and increases representation of the focal variables used to choose sites.

The general outline of SiteSelectR consists of five steps: (i) define the region in which to select sites as a spatial polygon; (ii) characterize the environmental conditions (e.g., elevation, land use, road density) within the region using geospatial data; (iii) calculate the dissimilarity in the environmental conditions among all sites using multivariate Mahalanobis distance – the region can be filtered to only accessible locations (e.g., public land); (iv) select a starting location (arbitrary) and select the most dissimilar site to the starting location; (v) recalculate the dissimilarity among all non-selected cells to the suite of conditions represented in the network. Iterate until the number of desired sampling sites. 
```{r setup}
library(SiteSelectR)
```
The SiteSelectR library works with the terra (Hijmans et al. 2025) and sf (Pebesma et al. 2025) packages Both are used in SiteSelectR to handle raster data (terra) and vectorized shapefiles (sf). 

In this document we show the use of site_select(), which is the main function of SiteSelectR. Then, we detail the different functions wrapped inside site_select() that maximizes representativeness of the field sites it chooses. For the purpose of this project, representativeness is how much of raster data used to select cells/sites is represented by the resulting cell values. 

Before running site_select() we'll provide functions to prepare raster data for site_select() and create a sampling extent to choose sites - two necessary steps for running site_select(). First, we'll use terra to randomly generate 4 rasters from a normal distribution to use an example. Each raster is generated at a different spatial extent to demonstrate how SiteSelectR handles projections and data organization. 
```{r}
var1<-terra::rast(matrix(rnorm(900, mean=0, sd=1), 
                       nrow=30, ncol=30), crs="EPSG:4326")
var2<-terra::rast(matrix(rnorm(625, mean=0, sd=1), 
                       nrow=25, ncol=25), crs="EPSG:4326")
var3<-terra::rast(matrix(rnorm(400, mean=0, sd=1), 
                       nrow=20, ncol=20), crs="EPSG:4326")
var4<-terra::rast(matrix(rnorm(1500, mean=0, sd=1), 
                         nrow=750, ncol=750), crs="EPSG:4326")
plot(var3)
```
We'll create our sampling extent, which is the spatial limit that we'll contain site selection. The sampling extent can be a park or other boundary defined by a vectorized shapefile. If this is the case load in the shapefile using sf::st_read() and project it to your focal coordinate reference system (CRS) using st_transform() and st_crs(). However, if this type of boundary is unnecessary or unattainable for the user, SiteSelectR offers a function, sample_ext() which calculates a circle surrounding a central starting location for users to build their own focal area. The starting location might be chosen arbitrarily (i.e., random spatial sampling), or it can be a location that's easily accessible.

The sample_ext() function writes a vectorized point using sf::st_as_sf() from latitude/longitude coordinates then projects the point to the desired CRS using an EPSG code. We use the World Geodetic System (WGS) 84 CRS since it's commonly used for GIS analysis and visualization. The sample extent uses a radius value (in meters) supplied by the user in sf::st_buffer() and uses the Mercator CRS projection which is conformable to a precise geospatial cylindrical calculation surrounding a single central point supplied by the user. Once calculated, the function converts the sampling extent to the supplied CRS. In this example we use 750,000 meters (m) and the coordinates 10.0, 12.0. 
```{r}
sampling_extent<-sample_ext(x=10.0, y=12.0, 750000, crs=4326)
head(sampling_extent)
```
Let's visualize the sampling extent on one of the simulated rasters. We arbitrarily chose the first raster to visualize our sampling extent. 
```{r}
plot(var1) ; lines(sampling_extent, col='red')
```
site_select() only chooses cells within the sample extent. site_select() uses terra::mask() in a few wrapped functions to ensure any cells outside the sampling extent are exluded - we demonstrate what this looks like here: 
```{r}
var1_mask<-terra::trim(terra::mask(var1, sampling_extent))
plot(var1_mask)
```
Use rast_collect() to take a set of rasters and place them in a usable list. rast_collect() takes any number of raster layers and combines them using list().
```{r}
r_list<-rast_collect(var1, var2, var3, var4)
head(r_list)
```
Next, we'll use rast_match() to resample the raster list to the same extent, resolution, and CRS - done using terra::project() and sf::st_transform(). This step is necessary to calculate a distance matrix that's spatially the same as our raster data. The raster list is projected to match a reference raster - we've arbitrarily chosen the fourth raster layer to be our reference. The extent should be the sampling extent either calculated using sample_ext() or an existing vectorized polygon. 
```{r}
r_resampled<-rast_match(raster_list = r_list, 
                        reference_layer = r_list[[4]], 
                        extent = sampling_extent)
head(r_resampled)
```
As a final data preparation step, SiteSelectR uses rast_stack() to convert the resampled list of rasters into a single georeferenced object, a raster stack, using terra::rast(). Then, rast_stack() harmonizes the raster object's extent and masks the stack to our sampling extent using terra::mask(). The extent and resolution of the rasters are harmonized often in SiteSelectR to ensure that the  Mahalanobis calculation matches their geospatial information exactly. If the rasters are not projected to the exact same extent and resolution for each data preparation step, the distance values will be spatially skewed and the resulting sites of SiteSelectR defunct. 
```{r}
r_stack<-rast_stack(r_resampled, sampling_extent)
names(r_stack)<-c("Raster 1", "Raster 2", "Raster 3", "Raster4")

plot(r_stack)

```
We now have a sampling extent and a stack of rasters - everything we need to run site_select(). We run this below, and then detail the components of the function afterwards. 

site_select() uses latitude and longitude values for the starting location. This can be the location use in the sample_ext() or something else. 'num' refers to number of additional sites to choose in additon to (1) the starting location (lat & lon), and (2) the first site chosen - both are calculated outside new_distance(), which is a function wrapped inside site_Select() that we detail more below. For now, that we'll always get 2 more sites than we put into site_select(). Here, we set num as 8, meaning our total count will be 10 sites. 
```{r}
sites<-site_select(long=12, lat=10, num=15, stack=r_stack)
print(sites)
```
We can visualize our sites using the plot_sites() function which takes the cells we generated and uses terra::XYFromcell() to transform them into coordinates. Then, plot_sites() plots the coordinates on a map of raster layer. plot_sites() returns a list of site coordinates.

```{r}
sites_coord<-plot_sites(cells=sites, stack=r_stack)
```
```{r}
print(sites_coord)
```
Let's break down site_select() and look at the functions that allowed us to produce this list of field sites. First, site_select() converts the raster stack into a matrix using as.matrix().Then, the first layer of the raster stack is set as an empty reference layer object  which will be used as a spatial reference for choosing cells and converting coordinates. 
```{r}
rast_data<-as.matrix(r_stack)
ref_layer<-r_stack[[1]]
values(ref_layer)<-NA
```
We can build the distance matrix using Mahalanobis distance calculation on the raster stack matrix. Mahalanobis is used because it's well equipped for multivariate environmental data that exhibits correlated features. The Mahalanobis distance is calculated as:

d^2 = (x_{ijk} - μ)' Σ⁻¹(x_{ijk} - μ)

where k is vector of environmental values at a given latitude(i) and longitude(j), μ is the mean vector across the raster layers – transposed to align with the inverse covariance matrix, Σ⁻¹ is the inverse of the covariance matrix, which adjusts for correlations among focal variables. This results in a scalar metric at each cell that reflects multivariate dissimilarity across all raster layers at once (Etherington, 2021; De Maesschalck, Jouan-Rimbaud, and Massart, 2000).

We created the covariance matrix using the cov() function, allowing for missing values by setting 'use' to pairwise.complete.obs – this is essential since NA values contribute to the spatial integrity of the raster data and cannot be removed. We added a diagonal of small values (0.01) to regularize the matrix which helps limit singularity issues between colinear data. We calculated the center mean for the Mahalanobis calculation using colMeans() on the raster data, ignoring NA values. We used the mahalanobis() function with a matrix of the raster stack data, center mean, and covariance matrix to produce the initial distance matrix. SiteSelectR uses as.numeric() as a final step to ensure the product is a plain numeric vector.

We wrap this process in the function, rast_distance(), which measures the dissimilarity between one cell to another relative to the mean distribution of the raster stack. 
```{r}
distance_matrix<-rast_distance(data=rast_data)
summary(distance_matrix)
```
We use heat_distance() to visualize the dissimilarity in our matrix. The highest dissimilar area will be chosen first. heat_distance() maps an empty raster of NA values, and then inserts the distance matrix values into the empty cells. 
```{r}
heat_map<-heat_distance(stack=distance_matrix)
plot(heat_map)
```
We now have a matrix of distance values, where each cell corresponds to a geographic location and its environmental variability. Next, we'll use max_cell() to choose the most dissimilar cell in the distance matrix. This will be our first site chosen based on the initial dissimilarity in the sampling extent.

```{r}
max_site<-max_cell(distance_matrix = distance_matrix)
print(max_site)
```
site_select() combines the two cells (the starting location and max distance site) into a usable format using cell_combine() and visualize them on our dissimilarity map (e.g., heat_distance() function). To wrap the coordinates used to build the sampling approach into a usable object we use cell_wrap() which converts our starting location into a set of georeferenced coordinates and uses terra::cellFromXY() into a single cell value that corresponds to our sampling extent. 
```{r}
cell1<-cell_wrap(x=12, y=10, raster_stack = r_stack)
cells<-cell_combine(max_site, cell1)
print(cells)
```
We can quick view these cells below. Notice the starting location was used to create the sampling extent and is in the center, and the max distance site is placed in the cell with the highest dissimilarity. 
```{r}
cells_xy<-terra::xyFromCell(r_stack[[1]], cells)
plot(heat_map); points(cells_xy, col='red', pch=19)
```
These 2 sites are the basis for automating the remaining code in the site_select() function, hence why we need to account for these sites when we input the number of desired sites argument ('num') into the function. The first Mahalanobis calculation represents the dissimilarity across the entire environmental gradient - this is represented by the product of max_cell(). Every following site selection must account for the environmental variation minus the cells that were already chosen. An approach that we include in the new_distance() function wrapped inside site_select(). 

new_distance() uses the Mahalanobis equation - same as rast_distance(). While rast_distance() measures the variability in one dataset (our raster data), new_distance() creates two datasets - a baseline group which is the raster data used in the rast_distance() function and the comparison group, which is the baseline group with the values of chosen cells replaced with NA values. By replacing the chosen cells with NA values, we preserve the spatial integrity of the raster data but remove those cells from consideration. Then, instead of calculating the dissimilarity between every cell to each other, new_distance() measures the dissimilarity between every cell not chosen yet (comparison group) to every possible cell (baseline group) simultaneously. 
```{r}
new_matrix<-new_distance(rast_data, cells)
summary(new_matrix)
```
Now choose another site with max_cell() on our newly constructed distance matrix. 
We combine this with our existing list of cells and plot. 
```{r}
third_site<-max_cell(new_matrix)
cells<-cell_combine(cells, third_site)
print(cells)
```
```{r}
new_sites<-plot_sites(cells, r_stack)
```
Let's automate this approach in a for-loop, which is what site_select() does when we pass the 'num' argument. We choose 15 iterations, meaning we'll get 17 sites including the initial two sites. Each iteration, site_select() updates the cells list, thereby ensuring an updated distance matrix.
```{r}
num=15
for (i in 1:num) {
  n <- new_distance(rast_data, cells)
  pt <- max_cell(n)
  cells <- cell_combine(cells, pt)  
}

print(cells)
```

```{r}
final_sites<-plot_sites(cells, r_stack)
```
We can use site_val() to pull the cell values from the original raster data for each field site. site_val() uses the terra::extract() function and converts the product to a matrix which can be used to assess how well site_select() chose sites that represent the environmental variability of our data.

```{r}
cell_values<-site_val(final_sites, r_stack)
print(cell_values)
```


With these values let's check if our sites represent the full variation across our raster data. To do this, we've developed two tools which might be helpful for users to assess their final output. (1) To view the colinearity between variables at site locations, vis_cor()  creates a correlation plot that shows the correlation between raster values at each site. This might useful to view how seperated correlated variables are at site locations (e.g., urban cover, population density, and temperature).
```{r}
cor_plot<-vis_cor(cell_values)
```
Each cell is a site, and at each site the value is either colinear (red) or noncolinear (blue) between the corresponding variables. vis_cor() returns a table of correlations as well. Using vis_cor() might be favorable so users know which variables can and can't be statistically separated a priori. 

(2) vis_rep() takes the site values from site_val() and plots two raster layers to compare where in their distribution our site values lie. The goal of SiteSelectR is to represent environmental space evenly during site selection - especially covering the maximum and minimum values (i.e., avoiding mean-driven sites) so it's favorable to have sites that occupy the entirety of each layer (CITE). 
```{r}
vis_rep(values=cell_values, stack=r_stack, layer1 = 1, 
        layer2 = 2)
```
